### 2019-03-07 paper/sex_selection_spacing-ver2.tex
Need earlier highlight of main questions

Birth spacing has long served as a measure of son preference, with strong
son preference typically associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination has the potential to
fundamentally change the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion automatically increases the interval between births by six 
months to a year.%
\footnote{
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Longer birth intervals may, therefore, cease to be an indication of declining son
preference, but could instead come about precisely because strong son preference makes
couples more likely to use sex selection.
% To further complicate matters, families with son preference who---for one reason or 
% another---do not use prenatal sex selection may continue to have shorter birth spacing 
% than those with a lower preference for sons.

@@

Even the effects of access to contraceptive on spacing are unclear.
On the one hand, access to contraceptives allows women to avoid too short spacing between 
birth, which would increase birth spacing.
On the other hand, increased reliability of access and effectiveness of contraceptives can 
lead to shorter intervals between births  \citep{Keyfitz1971,Heckman1976}.
With less reliable contraception parents choose a higher level of contraception, which 
results in longer spacing, to avoid having too many children by accident.
However, as contraception becomes more effective parents can more easily avoid future 
births, allowing them to reduce the spacing between births without having to worry about 
overshooting their preferred number of children.
This idea may explain why better-educated women have shorter spacing than less educated 
women in some instances but not in others.
These two counteracting effects may explain why finding statistically significant effects 
of contraception use on birth spacing is difficult \citep{Yeakey2009}.%
\footnote{
There is evidence in Matlab that the effect of son preference on birth spacing is
stronger in areas with better access to family planning \citep{Rahman1993}.
Other evidence, however, points to families ability to time birth, even in the absence
of modern contraceptives \citep{Jayachandran2011,Alam2018}.
}

TK spacing and decline in fertility (with and without sex selection)

Despite the increasing use of sex selection, the extensive use of spacing to measure son 
preference, and the important impacts of spacing on outcomes for both mother and children  
there has so far been no examination of what effect access to sex selection has had on 
spacing.

TK What questions I am asking

@@

% [Hypotheses / questions]

The overall hypothesis is that spacing patterns significantly changed 
as prenatal sex determination became available.
To address this, I examine three questions in order of importance.

First, did the spacing after girls increase relative to the spacing
after boys?
The underlying hypothesis is that in the absence of sex selection son 
preference leads to shorter spacing after the birth of a girl than after the 
birth of a boy, whereas son preference increases the spacing after the birth 
of a girl relative to after the birth of a boy when prenatal sex selection is 
available.
I still expect parents to \emph{try} to conceive earlier after the birth 
of a girl than after a boy, but the abortions will increase the 
duration between births, and may even lead to longer spacing after the
birth of girls than after boys.

Second, have the shapes of the survival curves changed?
The shapes of the survival curves provide information both on general
changes in birth spacing and whether parents are
still trying to conceive earlier after girls than boys.
I expect this to be the case if there is still a strong preference for
boys, and, as discussed above, this behavior may even become more pronounced
with access to sex selection.

Finally, does the sex ratio change within a spell?
There are two reasons why this is interesting.
First, if the sex ratio changes within a spell, relying on the observed sex
ratio at birth---as is standard practice in the literature---will not provide
correct estimates of the use of sex selection or the sex ratio at
the conclusion of childbearing.
Second, any changes in the sex ratio within a spell provides clues
to parents' preferences over the ideal spacing between births and the
distribution of son preference in the sample.

@@

>>>>

Birth spacing has long served as a measure of son preference, with strong
son preference typically associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination has the potential to
fundamentally change the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion automatically increases the interval between births by six 
months to a year.%
\footnote{
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Longer birth intervals may, therefore, cease to be an indication of declining son
preference, but could instead come about precisely because strong son preference makes
couples more likely to use sex selection.

@@

Even the effects of access to contraceptive and declining fertility on spacing are unclear.
On the one hand, access to contraceptives allows women to avoid too short spacing between 
birth, which would increase birth spacing.
On the other hand, increased reliability of access and effectiveness of contraceptives can 
lead to shorter intervals between births  \citep{Keyfitz1971,Heckman1976}.
With less reliable contraception parents choose a higher level of contraception, which 
results in longer spacing, to avoid having too many children by accident.
However, as contraception becomes more effective parents can more easily avoid future 
births, allowing them to reduce the spacing between births without having to worry about 
overshooting their preferred number of children.
This idea may explain why better-educated women have shorter spacing than less educated 
women in some instances but not in others.
These two counteracting effects may explain why finding statistically significant effects 
of contraception use on birth spacing is difficult \citep{Yeakey2009}.%
\footnote{
There is evidence in Matlab that the effect of son preference on birth spacing is
stronger in areas with better access to family planning \citep{Rahman1993}.
Other evidence, however, points to families ability to time birth, even in the absence
of modern contraceptives \citep{Jayachandran2011,Alam2018}.
}

% [Hypotheses / questions]
Despite the increasing use of sex selection, the important impacts of spacing on outcomes 
for both mother and children, and the extensive use of spacing to measure son 
preference in the literature, there has so far been no examination of what effect access 
to sex selection has had on spacing.
The overall hypothesis I examine here is that spacing patterns significantly changed 
as prenatal sex determination became available.
I address three questions in order of importance.

First, how did birth intervals change over time for different groups and how are those 
changes related to the likely use of sex selection?
The underlying hypothesis is that in the absence of sex selection son preference leads to 
shorter spacing the fewer boys a family has for a given parity, whereas son preference 
increases the spacing the fewer boys a family has when prenatal sex selection is available.
Families with son preference who---for one reason or another---do not use prenatal sex 
selection may continue to have shorter birth spacing.
Based on the prior literature on sex selection, I focus on how birth intervals have
changed for different education levels, whether urban or rural, and the sex composition of 
prior children.

Second, have the shapes of the survival curves changed?
The shapes of the survival curves provide information both on general changes in birth 
spacing and whether parents are still trying to conceive earlier after girls than boys.
I still expect parents to \emph{try} to conceive earlier after the birth 
of a girl than after a boy, but the abortions will increase the duration between births, 
and may even lead to longer average spacing after the birth of girls than after boys.
I expect this to be the case if there is still a strong preference for boys, and this 
behavior may even become more pronounced with access to sex selection.

Finally, how are the changes in birth intervals with increasing access to sex selection 
distributed geographically?
Given the size and cultural differences within India, it is unlikely that the effect of
increasing access to sex selection will have had the same effect throughout 
\citep{retherford03b}.
I, therefore, examine how birth intervals have changed across four broad regions.
[TK only some groups]

@@

@@

\subsection{Regional Differences\label{sec:regional}}
